Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2020-04-14 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex

\setkomavar{subject}{Revision of `JRSS-SC-Jan-20-0007'}
\setkomavar{signature}{Moritz Lang\\Corresponding Author}

\setkomavar{fromname}[Name]{Moritz Lang}
\setkomavar{fromemail}{Moritz.Lang@uibk.ac.at}

\setkomavar{fromaddress}{University of Innsbruck, Department of Statistics, Universit\"atsstrasse
15, 6020 Innsbruck, Austria}

\begin{letter}[fromemail]{To Prof. Peter Smith}
\opening{Dear Prof. Peter Smith}

We thank you, as well as the associate editor and both reviewers, for the
positive and constructive feedback regarding our manuscript ``\textbf{Circular
Regression Trees and Forests with an Application to Probabilistic Wind
Direction Forecasting}''.

\vspace{5mm}

We have carefully revised our manuscript according to given comments.
The most substantial changes in the manuscript are the following:

\begin{itemize}
\item We have fully revised the abstract in order to put a larger
focus on the application. In addition, we now use the term `application' 
instead of `case study' throughout the manuscript and have also renamed Sect.~4.

\item In the section computational details, we have additionally listed the needed
computing time of the different methods on a standard desktop computer.

\item We now discuss the work of Hara and Chellappa~(2017) and compare their distribution-based 
tree-building approach to our implementation of circular distributional trees within Sect.~1.1.

\item For the employed GLM implemented by Mulder and Klugkist~(2017), we now
explicitly state that the accommodation of circular covariates would be possible
but require an additional manual adjustment, e.g., an approximation by a finite
Fourier expansion (cf., Lund 1999).

\vspace{5mm}

On the following pages a point-to-point response to all comments will be given.
The attached manuscript highlights the changes in the text in blue color.

\closing{Your sincerely,}

\end{letter}

@@

\end{letter}

\newpage

\section{Editor: Prof. Peter Smith - Comments to the Author}

\begin{re}
Please can you rewrite the Summary to give the application more prominence.
Currently, the last sentence reads to me like an afterthought. Furthermore, I
would prefer you to use the term `application' rather than `case study'
throughout the paper, since in Applied Statistics papers the methodological
developments should be motivated by the application.
\end{re}

\vspace{-1em}
First of all we want to thank you for your valuable input.

We agree with your comment and have rewritten the abstract to put a larger
focus on the application. Throughout the manuscript we now use the term
`application' instead of `case study' and have also renamed Sect.~4
accordingly. \todo{Lisa: Revise abstract}

\vspace{0.5em}
\begin{re}
Page 2, line 46: `… structure can capture non-additive …'?
\end{re}

Thank you, we have rephrased the sentence according to your suggestion.

\vspace{0.5em}
\begin{re}
Page 2, lines 54 and 55: `… mean, whereas modelling the conditional …
distribution would also allow uncertainty in the forecast to be estimated.'?
\end{re}

We have rephrased the sentence accordingly.

\vspace{0.5em}
\begin{re}
Page 6, line 16 and Equation (2.6): bold the vector $\beta$.
\end{re}

Thank you for pointing that out, we have corrected that accordingly.

\vspace{0.5em}
\begin{re}
Page 6, lines 29 and 30: `In general, for additive models a proper model …'?
\end{re}

We have rephrased the sentence according to your suggestion.

\vspace{0.5em}
\begin{re}
Equations (3.1) to (3.4) and elsewhere: I find the use of multiple letters for a single superscript or subscript awkward and prefer to keep lower case letters for indices, so please consider replacing them, for example, T for tree and F for forest. Also for Equation (4.2), how about using parentheses rather than subscripts?
\end{re}

We have followed your suggestions and now use only single letters for super-
and subscripts in the main body of the manuscript. In the appendix, we keep the
notation as it is in order to be consistent with the reference Hothorn~\emph{et al.}~2006.

\vspace{0.5em}
\begin{re}
Equation (3.4): bold vector $z$.
\end{re}

Thank you for pointing that out.

\vspace{0.5em}
\begin{re}
Page 9, line 39: delete `respectively'.
\end{re}

We have corrected the sentence accordingly.

\vspace{0.5em}
\begin{re}
Page 15, line 23: replace `leveraged' with `used'?
\end{re}

We have rephrased the sentence accordingly.

Hothorn T, Hornik K, Zeileis A (2006). ``Unbiased Recursive Partitioning: A Conditional
Inference Framework.'' \emph{Journal of Computational and Graphical Statistics}, \textbf{15}\,(3), 651--674.
doi:10.1198/106186006x133933.

@@

Hothorn T, Hornik K, Zeileis A (2006). ``Unbiased Recursive Partitioning: A Conditional
Inference Framework.'' \emph{Journal of Computational and Graphical Statistics}, \textbf{15}\,(3), 651--674.
doi:10.1198/106186006x133933.

\newpage

\section{Associate Editor - Comments to the Author}

\begin{re}
Both referees agree this is a good paper, and have only a few suggestions for
changes, so I recommend minor revisions to address these. I have two short suggestions in addition to this:

Can you comment on the typical computational requirements of the various
methods used in your application, in terms of hardware required and length of
computing time.
\end{re}

First of all we want to thank you for your valuable input.

We agree with your comment that additional information on the calculation
requirements is valuable additional information for the user.  We have
therefore listed the needed computing time of the different methods on a
standard desktop computer within the section computational details.
\todo{Moritz: Run models}

\vspace{0.5em}
\begin{re}
Is it possible to include a link to computer code to reproduce your
analysis? This is not essential for publication but would be helpful for
readers who wish to perform similar analyses. The R package you have provided
for your method is very helpful in this regard.  
\end{re}

As we have used data from the Austrian air navigation service provider, which
we are not entitled to distribute, we are unfortunately not able to provide
self-contained computer code in order to reproduce our application. However, as you have already
pointed out the implementation of circular regression trees and forests is
provided in a freely available R-package with some minimal examples.
Independent of this publication, we are planning to provide an online
platform for the whole `partykit' project in the future.

@@

\newpage

\section{Referee: 1 - Comments to the Author}

\begin{re}
I found Circular Regression Trees and Forests to be a very nicely written
paper, with a meaningful contribution to the literature on regression models
for a circular response variable. 
\end{re}

\begin{re}
I have very few suggestions to the authors; but here are two:

On page 2, what is the reference for the phrase "Additionally, they
developed a combined ... version ...".
\end{re}

First of all we want to thank you for your review and your valuable input.

Thanks for pointing out that the citation here is not clearly assigned. We have therefore
listed the references below the statement again.

\vspace{0.5em}
\begin{re}
On page 11, regarding "As the implementation cannot handle circular
covariates ...": It could handle circular covariates in a fashion similar to
Lund (1999), using a rotational or reflective relationship, or modeling the
association with a trigonometric polynomial. Adding this to their comparison
model, or at least mentioning the possibility of doing so, would be more
appropriate than stating that circular covariates cannot be accommodated.
\end{re}

A very good remark. We have adapted the statement accordingly and now
explicitly state that the accommodation of circular covariates would be
possible but require an additional manual adjustment, e.g., an approximation by
a finite Fourier expansion.

@@

Lund UJ (1999). ``Least Circular Distance Regression for Directional Data.''
\emph{Journal of Applied Statistics}, \textbf{26}\,(6), 723--733.
doi:10.1080/02664769922160.

\newpage

\section{Referee: 2 - Comments to the Author}

\begin{re}
This is a well-written paper on using circular regression trees for analyzing a
specific type of circular data. The probabilistic approach is well-founded. The
experiments on wind direction data are well organized and the results are
thoroughly analyzed.
\end{re}

\begin{re}
The authors could address model order determination problem using the
Bayes Information Criterion (Schwartz, 1978, Annals of Statistics).  
\end{re}

First of all we want to thank you for your review and your valuable input.

Regarding your suggestion, when building
a circular distributional tree, the package \textbf{circtree} uses per default
\emph{p}-values from association tests both for selecting the split variable
and for a so-called pre-pruning strategy which stops growing the tree as soon as
no significant associations can be found in a given subgroup. This is the
default in both used unbiased recursive partitioning algorithms CTree
and MOB. For the latter, as discussed by Zeileis~\emph{et al.}~(2008), an
alternative, already implemented strategy  is to
use information criteria such as AIC (Akaike information criterion) or BIC
(Bayes information criterion) for post-pruning (after growing the initial tree). 
As we use neither MOB no post-pruning in the
presented application, we refrain from discussing this aspect any further in
the manuscript. However, we want to refer to the submitted manuscript by
Schlosser~\emph{et al.}~2019 which focuses on the advantages of different
unbiased recursive partitioning algorithm for tree building as suggested in the literature.

\vspace{0.5em}
\begin{re}
I would like to point out related works on circular regression models for
shape representation (Kashyap and Chellappa) and circular random forests for
vehicle direction estimation (Hara and Chellappa) given below. I feel there is
some overlap between this paper and the work of Hara and Chellappa on using
circular statistics for vehicle orientation estimation. The authors may include
some comparisons with Hara, Chellappa work.
\end{re}

Thank you very much for the very interesting and to us unfortunately so far unknown references.
We now review the work of Hara and Chellappa~(2017) and briefly compare their approach
to our implementation of circular distributional trees in Sect.~1.1.

@@

\vspace{0.5cm}
References

Hara K, Chellappa R (2017). ``Growing Regression Tree Forests by
Classification for Continuous Object Pose Estimation.'' \emph{International
Journal of Computer Vision}, \textbf{122}\,(2), 292--312.
doi:10.1007/s11263-016-0942-1.

Kashyap R, Chellappa R (1981). ``Stochastic Models for Closed Boundary
Analysis: Representation and Reconstruction.'' \emph{IEEE Transactions on
Information Theory}, \textbf{27}\,(5), 627--637. doi:10.1109/TIT.1981.1056390.

Schlosser L, Hothorn T, Zeileis A (2019). ``The Power of Unbiased Recursive
Partitioning: A Unifying View of CTree, MOB, and GUIDE.'' arXiv:1906.10179,
arXiv.org E-Print Archive.

Zeileis A, Hothorn T, Hornik K (2008). ``Model-Based Recursive Partitioning.''
\emph{Journal of Computational and Graphical Statistics}, \textbf{17}\,(2),
492--514. doi:10.1198/106186008x319331.

### 2020-04-14 pkg/circtree/inst/wind_paper/draft/JRSSC/rebuttal.tex

We thank you, as well as the associate editor and both reviewers, for the
positive and constructive feedback regarding our manuscript ``\textbf{Circular
Regression Trees and Forests with an Application to Probabilistic Wind
Direction Forecasting}''.

\vspace{5mm}

We have carefully revised our manuscript according to given comments.
The most substantial changes in the manuscript are the following:

\begin{itemize}
\item We have fully revised the `Abstract' in order to put a larger
focus on the application. In addition, we now use the term `application' 
instead of `case study' throughout the manuscript and have also renamed Sect.~4.

\item In the section `Computational details', we have additionally listed the needed
computing time of the different methods on a standard desktop computer.

@@

\item We now discuss the work of Hara and Chellappa~(2017) and compare their distribution-based 
tree-building approach to our implementation of circular distributional trees within Sect.~1.1.

\item Regarding the employed GLM implemented by Mulder and Klugkist~(2017), we now
explicitly state that the accommodation of circular covariates would be possible
but would require an additional manual adjustment, e.g., an approximation by a finite
Fourier expansion (cf., Lund 1999).

@@

\end{itemize}

\vspace{5mm}

On the following pages a point-to-point response to all comments will be given.
The attached manuscript highlights the changes in the main body in blue color,
the `Abstract' is fully revised but displayed in black color.

\closing{Yours sincerely,}

@@

\section{Editor: Prof. Peter Smith - Comments to the Author}

\begin{re}
Please can you rewrite the Summary to give the application more prominence.
Currently, the last sentence reads to me like an afterthought. Furthermore, I
would prefer you to use the term `application' rather than `case study'
throughout the paper, since in Applied Statistics papers the methodological
developments should be motivated by the application.
\end{re}

\vspace{-1em}
First of all we want to thank you for your valuable input.

We agree with your comment and have rewritten the `Abstract' to put a larger
focus on the application. Throughout the manuscript we now use the term
`application' instead of `case study' and have also renamed Sect.~4
accordingly.

@@

\vspace{0.5em}
\begin{re}
Equations (3.1) to (3.4) and elsewhere: I find the use of multiple letters for a single superscript or subscript awkward and prefer to keep lower case letters for indices, so please consider replacing them, for example, T for tree and F for forest. Also for Equation (4.2), how about using parentheses rather than subscripts?
\end{re}

We have followed your suggestions and now use only single letters for super-
and subscripts in the main body of the manuscript. In the appendix, we keep the
notation as it is in order to be consistent with the reference Hothorn~\emph{et al.}~2006.

\vspace{0.5em}
\begin{re}
Equation (3.4): bold vector $z$.
\end{re}

Thank you for pointing that out, the font style has been adapted.

@@

\section{Associate Editor - Comments to the Author}

\begin{re}
Both referees agree this is a good paper, and have only a few suggestions for
changes, so I recommend minor revisions to address these. I have two short suggestions in addition to this:

Can you comment on the typical computational requirements of the various
methods used in your application, in terms of hardware required and length of
computing time.
\end{re}

We first want to thank you for the positive feedback and your valuable input.

We agree with your comment that additional information on the calculation
requirements is valuable for the user. We have therefore listed the needed
computing time of the different methods on a standard desktop computer within
the section 'Computational details'.  \todo{Moritz: Run models}

@@

\vspace{0.5em}
\begin{re}
Is it possible to include a link to computer code to reproduce your
analysis? This is not essential for publication but would be helpful for
readers who wish to perform similar analyses. The R package you have provided
for your method is very helpful in this regard.  
\end{re}

As we have used data from the Austrian air navigation service provider, which
we are not entitled to distribute, we are unfortunately not able to provide
self-contained computer code in order to reproduce our application. However, as you have already
pointed out the implementation of circular regression trees and forests is
provided in a freely available R-package with some minimal examples.
Independent of this publication, we are planning to provide an updated online
platform for the whole `partykit' project in the future.

@@

\begin{re}
I found Circular Regression Trees and Forests to be a very nicely written
paper, with a meaningful contribution to the literature on regression models
for a circular response variable. 
\end{re}

\begin{re}
I have very few suggestions to the authors; but here are two:

On page 2, what is the reference for the phrase "Additionally, they
developed a combined ... version ...".
\end{re}

First of all we want to thank you for your positive and constructive feedback.

@@

Thanks for pointing out that the citation here is not clearly assigned. We have therefore
listed the references below the statement again.

\vspace{0.5em}
\begin{re}
On page 11, regarding "As the implementation cannot handle circular
covariates ...": It could handle circular covariates in a fashion similar to
Lund (1999), using a rotational or reflective relationship, or modeling the
association with a trigonometric polynomial. Adding this to their comparison
model, or at least mentioning the possibility of doing so, would be more
appropriate than stating that circular covariates cannot be accommodated.
\end{re}

A very good remark. We have adapted the statement accordingly and now
explicitly state that the accommodation of circular covariates would be
possible but would require an additional manual adjustment, e.g., an approximation by
a finite Fourier expansion.

@@

\section{Referee: 2 - Comments to the Author}

\begin{re}
This is a well-written paper on using circular regression trees for analyzing a
specific type of circular data. The probabilistic approach is well-founded. The
experiments on wind direction data are well organized and the results are
thoroughly analyzed.
\end{re}

\begin{re}
The authors could address model order determination problem using the
Bayes Information Criterion (Schwartz, 1978, Annals of Statistics).  
\end{re}

First of all we want to thank you for your positive feedback and valuable input.

Regarding your suggestion, when building a circular distributional tree, the
package \textbf{circtree} uses per default \emph{p}-values from association
tests both for selecting the split variable and for a so-called pre-pruning
strategy which stops growing the tree as soon as no significant associations
can be found in a given subgroup. This is the default in both available
unbiased recursive partitioning algorithms CTree and MOB. For the latter, as
discussed by Zeileis~\emph{et al.}~(2008), an alternative, already implemented
strategy  is to use information criteria such as AIC (Akaike information
criterion) or BIC (Bayes information criterion) for post-pruning (after growing
the initial tree).  As we use neither MOB nor post-pruning in the presented
application, we refrain from discussing this aspect any further in the
manuscript. However, we want to refer to the submitted manuscript by
Schlosser~\emph{et al.}~2019 which investigates the performance of different 
testing and pruning strategies employed in unbiased recursive partitioning algorithms.

### 2020-04-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update rebuttal and draft (Lisa Input and new Abstract)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3009 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4bc02d09fa6af89983ac9884ec756193c0a9a91a

\Abstract{
While circular data occur in a wide range of scientific fields, the methodology
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. Most of the existing methods are built on the
framework of generalized linear and additive models, which are often
challenging to optimize and interpret. Therefore, building on previous ideas
for trees modeling circular means, we suggest a distributional approach for
regression trees and random forests yielding probabilistic forecasts based on
the von~Mises distribution. The resulting tree-based models simplify the
estimation process by using the available covariates for partitioning the data
into sufficiently homogeneous subgroups so that a simple von~Mises distribution
without further covariates can be fitted to the circular response in each
subgroup. These circular regression trees are straightforward to interpret, can
capture nonlinear effects and interactions, and automatically select the
relevant covariates that are associated with either location and/or scale
changes in the von~Mises distribution. Combining an ensemble of circular
regression trees to a circular regression forest can regularize and smooth the
covariate effects. The new methods are evaluated in a case study on
probabilistic wind direction forecasting at two Austrian airports, considering
other common approaches as a benchmark.
}

@@

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already introduced a
circular regression tree algorithm where binary splits are made based on an
angular distance measure capturing node homogeneity. However, this only models
changes in the conditional mean, whereas modeling the conditional variance or
full probabilistic distribution would also allow uncertainty in the forecast to
be estimated \citep{Gneiting:2008}. 
%However, this only models
%changes in the conditional mean but not the conditional variance or full
%probabilistic distribution which would allow for also considering uncertainty
%in forecasts \citep{Gneiting:2008}. 
In the field of image based object direction estimation,
\citet{Hara+Chellappa:2017} presented a distribution-based tree
building algorithm by applying $k$-means clustering to the target space where
the number of clusters $k$ is adaptively determined by the Bayes
information criterion employing the von~Mises distribution. Despite their
distribution-based splitting approach to circular response variables, they
solely predict the mode of the distribution, thus, providing no full
probabilistic forecasts. Combining these previous ideas for circular trees with the
concept of distributional modeling, we introduce a fully distributional tree
approach that considers both splits in all distribution parameters and yields a
full probabilistic predictive model, as proposed by \citet{Gneiting:2008}. In
addition, as a natural extension, ensembles or forests of circular regression
trees are presented in this study, showing that these can further improve
predictive power.

@@

Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with the
empirical (gray) and fitted von~Mises~(red) wind direction distribution in each
terminal node. Based on the fitted location parameters~$\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2)~Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3)~Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and~13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more comprehensive application of circular regression
trees and forests on probabilistic wind direction forecasting is presented in
Section~\ref{sec:wind} for Innsbruck Airport and Vienna International Airport,
together with a comparison with commonly-used alternative approaches.

@@

% -------------------------------------------------------------------
% Application
% -------------------------------------------------------------------
\section{Application: Probabilistic wind direction forecasting}
\label{sec:wind}
As motivated in Section~\ref{sec:introduction}, accurate forecasts of wind
directions are of great importance for risk management in various fields such
as agriculture, energy production, or aviation. For example, in order to direct
airplanes to a safe landing, precise knowledge of wind direction for the next
hour(s) at the respective airport is highly desirable and adequate prediction
methods are required. This section exemplifies the use of circular regression
trees and forests with wind direction forecasts for two Austrian airports --
one in flat terrain, the other one in mountainous terrain. The results are
benchmarked against alternative probabilistic forecasting methods. The study is
based on $+1$\,h and $+3$\,h forecasts employing lagged observations in the
vicinity of the airports as possible predictor variables.

\subsection{Data}
%%
\begin{figure}[t]
  \setkeys{Gin}{width=1.0\textwidth}
  \centering
  \includegraphics{_plot_circforest_stationmap_combined.pdf}
  \caption{Overview of the study area for Innsbruck Airport (left) and Vienna International Airport (right). 
    For Innsbruck, $4$~stations at the airport and $6$~stations along the
    intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
    and $13$~stations within the vicinity of approximately 30\,km are used.
    Elevation data are obtained from the TandDEM-X digital elevation model with a 
    horizontal resolution of $90$~m \citep{Wessel:2018}.}
  \label{fig:overview} 
\end{figure}
%%
\label{sec:wind:data}
The circular response variable considered in this application is a
$10$\,min-average of wind direction measurements at Innsbruck Airport (INN) and
Vienna International Airport (VIE) on an hourly temporal resolution. Temporal
information and $1$-hourly resolved $10$\,min mean observations of various
meteorological quantities are used as predictor variables, including wind
direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within its vicinity of approximately 30\,km are used.
Figure~\ref{fig:overview} provides a topographical overview of the airports and
their surrounding areas with the station sites employed in this study. In
addition, we use derived quantities such as 3-hourly means, minima and maxima,
as well as 1-and 3-hourly temporal changes and spatial differences towards the
airport of the respective quantities. An overview of the employed data sets can
be found in Table~\ref{tab:data}.

The data used in this study consists of five subsequent years from January~2014
to December~2018. After first eliminating predictor variables with more than
5\% missing values and then time points with any missing observations, the data
set consists of 41\,979~time points and $260$~covariates for Innsbruck, and of
38\,985~time points and $494$~covariates for Vienna.

@@

\subsection{Models and evaluation}
\label{sec:wind:models}
For a fair evaluation of circular regression trees and forests, and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches used in the meteorological field, while the third is a
state-of-the-art GLM-type model to forecast circular response variables.

@@

The performance of the novel circular regression trees and forests is assessed
in an application on shortterm probabilistic wind direction forecasting at two
airports with different site characteristics. As benchmark models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM-type model are evaluated based on proper scoring rules. In
summary, the circular regression trees and forests have the highest predictive
performance in this setting. For cases without changes in the wind regime,
lagged response values provide already highly skillful estimates leading to a
good performance of the persistency model as observed for shortterm wind
direction forecasts in this study. While in these cases the trees and forests
also benefit from the highly informative lagged response, they can compensate
for a lower information of this covariate by incorporating other quantities and
possible interactions of these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods.

@@

\section*{Computational details}
The corresponding implementation of the proposed methodology for circular
regression trees and forests is provided in the \textsf{R} package
\textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The package is
based on the \textbf{disttree} package
(version~\Sexpr{packageVersion("disttree")}) which applies the main tree
building functions from the \textbf{partykit} package
(version~\Sexpr{packageVersion("partykit")}). All three packages are available
on \textsf{R}-Forge at {\url{https://R-Forge.R-project.org/projects/partykit/}}.

@@

\section{Tree algorithm}
\label{app:algorithm}
This section provides a more detailed overview on the permutation-test-based
CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006}, specifically for circular
data as applied for building circular regression trees and forests presented in
this application. An alternative tree-building framework is provided by the MOB
algorithm, which is based on M-fluctuation tests \citep[see][for more
details]{Zeileis+Hothorn+Hornik:2008}.

>>>>

\Abstract{
%SUBMITTED VERSION:
%While circular data occur in a wide range of scientific fields, the methodology
%for distributional modeling and probabilistic forecasting of circular response
%variables is rather limited. Most of the existing methods are built on the
%framework of generalized linear and additive models, which are often
%challenging to optimize and interpret. Therefore, building on previous ideas
%for trees modeling circular means, we suggest a distributional approach for
%regression trees and random forests yielding probabilistic forecasts based on
%the von~Mises distribution. The resulting tree-based models simplify the
%estimation process by using the available covariates for partitioning the data
%into sufficiently homogeneous subgroups so that a simple von~Mises distribution
%without further covariates can be fitted to the circular response in each
%subgroup. These circular regression trees are straightforward to interpret, can
%capture nonlinear effects and interactions, and automatically select the
%relevant covariates that are associated with either location and/or scale
%changes in the von~Mises distribution. Combining an ensemble of circular
%regression trees to a circular regression forest can regularize and smooth the
%covariate effects. The new methods are evaluated in a case study on
%probabilistic wind direction forecasting at two Austrian airports, considering
%other common approaches as a benchmark.
%
%NEW VERSION:
Circular data occur in a wide range of scientific fields, but the methodology
for distributional modeling and probabilistic forecasting of circular response
variables is rather limited. While most of the existing methods are built on
the framework of generalized linear and additive models, which are often
challenging to optimize and interpret, we propose a distributional approach for
regression trees yielding probabilistic forecasts based on the von~Mises
distribution. The resulting tree-based models use the available covariates only
for partitioning the data into sufficiently homogeneous subgroups, allowing for
a simple circular distributional fit in each subgroup. These circular
regression trees are straightforward to interpret, can capture nonlinear
effects and interactions, and automatically select the relevant covariates that
are associated with changes in any distribution parameter. Ensembles or
forests of such circular trees can further improve the forecasts by
regularizing and stabilizing the model. The newly developed methodology is
illustrated in an application on short-term probabilistic wind direction
forecasting for Innsbruck and Vienna Airport in order to direct airplanes to a
safe landing. The predictive skill of the novel approaches is benchmarked with
an additive regression model plus a persistency and a climatology model,
employing the circular continuous ranked probability score.}

@@

This novel approach to circular regression trees and forests complements the
literature on tree-based circular modeling. \cite{Lund:2002} already introduced a
circular regression tree algorithm where binary splits are made based on an
angular distance measure capturing node homogeneity. However, this only models
changes in the conditional mean, whereas modeling the conditional variance or
full probabilistic distribution would also allow uncertainty in the forecast to
be estimated \citep{Gneiting:2008}. 
%However, this only models
%changes in the conditional mean but not the conditional variance or full
%probabilistic distribution which would allow for also considering uncertainty
%in forecasts \citep{Gneiting:2008}. 
In the field of image based object direction estimation,
\citet{Hara+Chellappa:2017} presented a distribution-based tree-building 
algorithm by applying $k$-means clustering to the target space where
the number of clusters $k$ is adaptively determined by the Bayes
information criterion employing the von~Mises distribution. Despite their
distribution-based splitting approach to circular response variables, they
solely predict the mode of the distribution, thus, providing no full
probabilistic forecasts. Combining these previous ideas for circular trees with the
concept of distributional modeling, we introduce a fully distributional tree
approach that not only considers splits in all distribution parameters, but also yields a
full probabilistic predictive model as proposed by \citet{Gneiting:2008}. In
addition, as a natural extension, ensembles or forests of circular regression
trees are presented in this study, showing that these can further improve
predictive power.

@@

Figure~\ref{fig:tree_ibk} illustrates the resulting tree along with the
empirical (gray) and fitted von~Mises~(red) wind direction distribution in each
terminal node. Based on the fitted location parameters~$\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2)~Downslope winds blowing across the Alpine crest along
the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3)~Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and~13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind
regimes and are consequently associated with a very low estimated concentration
parameter~$\hat{\kappa}$, i.e., a high estimated variance. In terms of
covariates, the lagged wind direction (``persistence'') is mostly responsible
for distinguishing the broad range of wind regimes listed above while the
pressure gradients and wind speed separate the data into subgroups with high
vs.\ low precision. A more comprehensive application of circular regression
trees and forests on probabilistic wind direction forecasting is presented in
Section~\ref{sec:wind} for Innsbruck Airport and Vienna International Airport,
together with a comparison to commonly-used alternative approaches.

@@

% -------------------------------------------------------------------
% Application
% -------------------------------------------------------------------
\section{Application: Probabilistic wind direction forecasting}
\label{sec:wind}
As motivated in Section~\ref{sec:introduction}, accurate forecasts of wind
directions are of great importance for risk management in various fields such
as agriculture, energy production, or aviation. For example, in order to direct
airplanes to a safe landing, precise knowledge of wind direction for the next
hour(s) at the respective airport is highly desirable and adequate prediction
methods are required. This section exemplifies the use of circular regression
trees and forests with wind direction forecasts for two Austrian airports --
one in flat terrain, the other one in mountainous terrain. The results are
benchmarked against alternative probabilistic forecasting methods. The application is
based on $+1$\,h and $+3$\,h forecasts employing lagged observations in the
vicinity of the airports as possible predictor variables.

\subsection{Data}
%%
\begin{figure}[t]
  \setkeys{Gin}{width=1.0\textwidth}
  \centering
  \includegraphics{_plot_circforest_stationmap_combined.pdf}
  \caption{Overview of the study area for Innsbruck Airport (left) and Vienna International Airport (right). 
    For Innsbruck, $4$~stations at the airport and $6$~stations along the
    intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
    and $13$~stations within the vicinity of approximately 30\,km are used.
    Elevation data are obtained from the TandDEM-X digital elevation model with a 
    horizontal resolution of $90$~m \citep{Wessel:2018}.}
  \label{fig:overview} 
\end{figure}
%%
\label{sec:wind:data}
The circular response variable considered in this application is a
$10$\,min-average of wind direction measurements at Innsbruck Airport (INN) and
Vienna International Airport (VIE) on an hourly temporal resolution. Temporal
information and $1$-hourly resolved $10$\,min mean observations of various
meteorological quantities are used as predictor variables, including wind
direction, wind speed, temperature, air pressure and humidity, all lagged by
one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
their vicinities. For Innsbruck, measurements at the airport and along the
intersecting valleys are used, whereas, for Vienna, measurements at the airport
and within its vicinity of approximately 30\,km are used.
Figure~\ref{fig:overview} provides a topographical overview of the airports and
their surrounding areas with the station sites employed in the application. In
addition, we use derived quantities such as 3-hourly means, minima and maxima,
as well as 1-and 3-hourly temporal changes and spatial differences towards the
airport of the respective quantities. An overview of the employed data sets can
be found in Table~\ref{tab:data}.

The data used in the application consists of five subsequent years from January~2014
to December~2018. After first eliminating predictor variables with more than
5\% missing values and then time points with any missing observations, the data
set consists of 41\,979~time points and $260$~covariates for Innsbruck, and of
38\,985~time points and $494$~covariates for Vienna.

@@

\subsection{Models and evaluation}
\label{sec:wind:models}
For a fair evaluation of circular regression trees and forests, and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in the
application on probabilistic wind direction forecasting. Two of them are based
on existing approaches used in the meteorological field, while the third is a
state-of-the-art GLM-type model to forecast circular response variables.

@@

The performance of the novel circular regression trees and forests is assessed
in an application on shortterm probabilistic wind direction forecasting at two
airports with different site characteristics. As benchmark models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM-type model are evaluated based on proper scoring rules. In
summary, the circular regression trees and forests have the highest predictive
performance in this setting. For cases without changes in the wind regime,
lagged response values provide already highly skillful estimates leading to a
good performance of the persistency model as observed for shortterm wind
direction forecasts in the application. While in these cases the trees and forests
also benefit from the highly informative lagged response, they can compensate
for a lower information of this covariate by incorporating other quantities and
possible interactions of these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods.

@@

\section*{Computational details}
The corresponding implementation of the proposed methodology for circular
regression trees and forests is provided in the \textsf{R} package
\textbf{circtree} (version~\Sexpr{packageVersion("circtree")}). The package is
based on the \textbf{disttree} package
(version~\Sexpr{packageVersion("disttree")}) which applies the main tree-building 
functions from the \textbf{partykit} package
(version~\Sexpr{packageVersion("partykit")}). All three packages are available
on \textsf{R}-Forge at {\url{https://R-Forge.R-project.org/projects/partykit/}}.

@@

\section{Tree algorithm}
\label{app:algorithm}
This section provides a more detailed overview on the permutation-test-based
CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006}, specifically for circular
data as applied for building circular regression trees and forests presented in
this study. An alternative tree-building framework is provided by the MOB
algorithm, which is based on M-fluctuation tests \citep[see][for more
details]{Zeileis+Hothorn+Hornik:2008}.
